---
title: "Class17_VaccinationMiniProject"
author: "Caitriona Brennan"
Date: "11-24-2021"
output:
  pdf_document:
    latex_engine: xelatex
---


#Background

In this Thanksgiving class when many of our classmates are traveling let's have a look at COVID-19 vaccination rates around the States.

We get vaccination rate data from CA.GOV.


#Import Data

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

>Q. How many entries do we have?

```{r}
nrow(vax)
```


We can use the **skimr** package and the 'skim()' function to get a quick overview of structure of this dataset.
IF we only want to use Skimr once we can call it like this rather than library(skimr)

```{r}
skimr::skim(vax)
```
#Notice that one of these is a date column. Working with time and dates get's annoying quickly.

```{r}
library(lubridate)

today()
```

>Q. How many days since the first entry in the dataset?

```{r}
vax$as_of_date[1]
```


This will not work because our data column was read as a character 
```{r}
#today() - vax$as_of_date[1]
```



```{r}
d <- ymd(vax$as_of_date)
```

```{r}
today() - d[1]
```

I will make the 'as_of_date' column date format...

```{r}
#vax$as_of_date <- ymd(vax$as_of_date)
```

>Q. When was the dataset last updated? What is the last date in this dataset?How many days since the last update?

```{r}
#today() - vax$as_of_date[nrow(vax)]
```

>Q. How many days does this dataset span?

```{r}
#vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

>Q. How many different ZIP codes are recorded in this dataset?

```{r}
zipcodes <- unique(vax[,2])
length(zipcodes)
```
```{r}
library(zipcodeR)
```



##Focus in on San Diego County

We want to subset the full CA vax data down to just San Diego. 

```{r}
inds <- vax$county == "San Diego"
nrow(vax[inds,])
```

Subsetting can get tedious and complicated quickly when you have multiple things we want to subset by


```{r}
library(dplyr)
```

We will use the "filter()" function to do our subsetting from now on

WE want to focus in on san diego county

```{r}
sd <- filter(vax, county == "San Diego")
nrow(sd)
```

To do more complicated subsetting...

```{r}
sd.20 <- filter(vax, county=="San Diego", 
       age5_plus_population > 20000)

nrow(sd.20)
```


```{r}
head(sd.20)
```

```{r}
length(unique(sd.20[,2]))
```

```{r}
sd.now <- filter(vax, county=="San Diego", 
       as_of_date=="2021-11-23")

nrow(sd.now)
```


```{r}
summary(sd.now$percent_of_population_fully_vaccinated)
```


>Q. Make a histogram of these values:

R based histogram

```{r}
hist(sd.now$percent_of_population_fully_vaccinated)
```

This plot above is going to be susceptible to being skewed by ZIP code areas with small populations. These will have big effects for just a small number of unvaxed folks....

>Q. what is the population of the 92037 ZIP code area?

>Q. what is the average vaccination value for this UCSD/La Jolla ZIP code area?

```{r}
lj <- filter(sd.now, sd.now$zip_code_tabulation_area==92037)
lj$age5_plus_population
```

```{r}
lj$percent_of_population_fully_vaccinated
```

```{r}
Hillcrest <- filter(sd.now, sd.now$zip_code_tabulation_area==92103)
Hillcrest$age5_plus_population
Hillcrest
```
```{r}
Hillcrest$percent_of_population_fully_vaccinated
```

Time series of vaccination rate for a given ZIP code area.
Start with 92037


```{r}
hillcrest <- filter(vax, vax$zip_code_tabulation_area==92103)
head(hillcrest)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(hillcrest) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

Let's make this plot for all San Diego county ZIP code areas that have a population as least as large as 92037. 


```{r}
sd.36 <- filter(vax, county=="San Diego",
             age5_plus_population > 36144)
head(sd.36)
```

>How many ZIP code areas in San Diego county have a population larger than 92037?


```{r}
length(unique(sd.36$zip_code_tabulation_area))
```

Let's make the plot



```{r}
ggplot(sd.36) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated,
      group=zip_code_tabulation_area) +
  geom_line() +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

>Q. Make a plot like this for all ZIP codes in CA that have a population at least as large as La Jolla (>31644)


```{r}
ca.36 <- filter(vax,
                age5_plus_population > 36144)
head(ca.36)
```

How many zipcode areas?

```{r}
length(unique(ca.36$zip_code_tabulation_area))
```

```{r}
ggplot(ca.36) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated,
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated") +
  geom_hline(yintercept = 0.67, color="red")


```

>Q. What is the mean accross the state for these 36K + population areas?

```{r}
ca.now <- filter(ca.36, as_of_date=="2021-11-23")
summary(ca.now$percent_of_population_fully_vaccinated)

```







