---
title: 'Class_09_mini_project'
author: 'Caitriona Brennan'
output: github_document
---


```{r}
#read.csv("WisconsinCancer.csv")
```
# Save your input data file into your Project directory
```{r}
fna.data <- "WisconsinCancer.csv"
```
# Complete the following code to input the data and store as wisc.df
```{r}
wisc.df <- read.csv(fna.data, row.names=1)
```

```{r}
View(fna.data)
```
#Q1. How many observations are in this dataset?
Answer - 569 observations
```{r}
#wisc.df
```

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```

```{r}
# Create diagnosis vector for later 
diagnosis <- as.factor(wisc.df$diagnosis)
diagnosis
```
#Q2. How many of the observations have a malignant diagnosis?
#Answer - 212 
```{r}
table(diagnosis)
```
```{r}
summary(wisc.data)
```
#Q3. How many variables/features in the data are suffixed with _mean?
#Answer - 10

```{r}
#needed to take just the col names
#Then find a pattern in col names that ends with _mean
#then use length to count these instances
colnames(wisc.df)
length(grep("_mean", colnames(wisc.df)))
```

# Check column means and standard deviations
```{r}
colMeans(wisc.data)
apply(wisc.data,2,sd)
```


```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
```

```{r}
summary(wisc.pr)
```
```{r}
#biplot(wisc.pr)
```
#Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
# Answer - 44.27%

#Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
#answer - 3

#Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
#Answer - 7

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

```{r}
biplot(wisc.pr)
```

#Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?
#Answer - difficult to understand, it is too compressed and saturated with data.


```{r}
# Scatter plot observations by components 1 and 2
plot( wisc.pr$x[,1:2] , col = diagnosis , 
     xlab = "PC1", ylab = "PC2")
```
```{r}
#Q8 Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
plot( wisc.pr$x[,c(1,3)] , col = diagnosis , 
     xlab = "PC1", ylab = "PC3")
```


```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
total.var <- sum(pr.var)
pve <- pr.var / total.var
pve

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

# Alternative scree plot of the same data, note data driven y-axis
```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```
```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```
 #Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? 
 #Answer - -0.26085376

```{r}
wisc.pr$rotation[,1]
```
#Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?
#Answer - 5  
```{r}
summary(wisc.pr)
```

```{r}
#need to call summary of our PCA and then assign it to a variable
#this summary has importance table. we get the sum of the 3 row of this table and see when it is less than 0.8.
#this returns 4 but we would beed at least 5 PCs to get over 0.8.
var <- summary(wisc.pr)
sum(var$importance[3,] < 0.8)
```
```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
```

```{r}
wisc.hclust <- hclust(data.dist, method = "complete" )
```

```{r}
plot(wisc.hclust)
abline(h = 19, col="red", lty=2)
```
#Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?
#Answer - 19

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 2, h = 19)
```


#Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
table(wisc.hclust.clusters, diagnosis)
```






#Section 5
Here we aim to combine our PCA results with clustering. Essentially, we are going to cluster in "PC space", that is cluster on the 'wisc.pr$x' results.


```{r}
summary(wisc.pr)
```

##CLuster my PCA results
I will use 4 PCs and 'hclust()' and 'dist()' as an input.

```{r}
wisc.pr.hlust <- hclust( dist(wisc.pr$x[,1:4]), method = "ward.D2")
```


```{r}
plot(wisc.pr.hlust)
abline(h=80, col = "red")
```

let's find our cluster membership vector by cutting this tree into k=2 groups

```{r}
grps <- cutree(wisc.pr.hlust, k=2)
table(grps)
```

Now let's compare to the expert M and B vector

```{r}
table(diagnosis)
```

we can do a cross-table by giving the 'table()' function two inputs
#the  B (6) below can be called false neg as they are in that clsuter that has mostly M

```{r}
table(grps, diagnosis)
```

**Accuracy** essentially how many did we get correct?
#nrow to get the total number of patients 
We got 90% accuracy 
```{r}
(165+351)/nrow(wisc.data)
```
#Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

**Sensitivity** refers to a test’s ability to correctly detect ill patients who do have the condition. In our example here the sensitivity is the total number of samples in the cluster identified as predominantly malignant (cancerous) divided by the total number of known malignant samples. In other words: TP/(TP+FN).

```{r}
(165)/((165)+(6))
```
**Specificity** relates to a test’s ability to correctly reject healthy patients without a condition. In our example specificity is the proportion of benign (not cancerous) samples in the cluster identified as predominantly benign that are known to be benign. In other words: TN/(TN+FN).


```{r}
351/(351+6)
```

# PREDICTION 
We will use the predict() function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

#Now add these new samples to our PCA plot
#pch = making points solid fill
cex - makes these points bigger size 3
text here labels the points 1 and 2 and makes these numbers white

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], labels=c(1,2), col="white")
```



