---
title: 'Class 10: Candy mini project'
author: 'Caitriona Brennan'
date: "10/29/2021"
output: github_document
---

```{r}
candy_file <- "candy.csv"

candy =  read.csv(candy_file, row.names=1)
head(candy)
```

#Q1. How many different candy types are in this dataset?
#Answer 85

```{r}
nrow(candy)
```

#Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy[,"fruity"])
```

```{r}
sum(candy[,"chocolate"])
```


#Q3. What is your favorite candy in the dataset and what is it’s winpercent value?


```{r}
candy["Almond Joy", ]$winpercent
```

Q4. What is the winpercent value for “Kit Kat”?
```{r}
candy["Kit Kat", ]$winpercent
```
Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent

```




#side-note: the skimr::skim() function

There is a useful skim() function in the skimr package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library(skimr)
skim(candy)
```

#QQ6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset? - Answer = winpercent

#Q7. What do you think a zero and one represent for the candy$chocolate column? Answer = Absent and present


#Q8. Plot a histogram of winpercent values


```{r}
hist(candy$winpercent)
```

#Q9. Is the distribution of winpercent values symmetrical? No

#Q10. Is the center of the distribution above or below 50%? Below

#Q11. On average is chocolate candy higher or lower ranked than fruit candy? Answer = Higher
#as logical to return which candy is chocolate
```{r}
as.logical(candy$chocolate)
```
#for all chocolate in candy what is the winpercent
```{r}
candy[as.logical(candy$chocolate),"winpercent"]
```
#get the average of these 
```{r}
chocolate <- candy[as.logical(candy$chocolate), "winpercent"]
mean(chocolate)
```
```{r}
fruity <- candy[as.logical(candy$fruity), "winpercent"]
mean(fruity)
```
## Q12. Is this difference statistically significant? ans = Yes

#it matters the order you put chocolate and candy. 
```{r}
t.test(chocolate, fruity)
```

#3. Overall Candy Rankings
Let’s use the base R order() function together with head() to sort the whole dataset by winpercent. Or if you have been getting into the tidyverse and the dplyr package you can use the arrange() function together with head() to do the same thing and answer the following questions:


Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

#we need to improve this to reorder the candy by the winpercent values
##Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent,reorder(rownames(candy), winpercent)) +
  geom_col()
```

#create colors for every candy type

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

##Q13. What are the five least liked candy types in this set?
```{r}
head(candy[order(candy$winpercent),], n=5)
```


##Q14. What are the top 5 all time favorite candy types out of this set?
```{r}
#candy %>% arrange(winpercent) %>% tail(5)
```





```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text()
```

#Q17. What is the worst ranked chocolate candy?
Answer - Nik L Nip
#Q18. What is the best ranked fruity candy?
Answer - Reeses peanut butter cup

#4. Taking a look at pricepercent
What about value for money? What is the the best candy for the least money? One way to get at this would be to make a plot of winpercent vs the pricepercent variable. The pricepercent variable records the percentile rank of the candy’s price against all the other candies in the dataset. Lower vales are less expensive and high values more expensive.

To this plot we will add text labels so we can more easily identify a given candy. There is a regular geom_label() that comes with ggplot2. However, as there are quite a few candys in our dataset lots of these labels will be overlapping and hard to read. To help with this we can use the geom_text_repel() function from the ggrepel package.




```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 10)
```
#Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?


```{r}
rownames(candy)
```

#change Õ in the rownames to a '. gsub = global substitute
```{r}
gsub("Õ", "'", rownames(candy))
```

```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord,c(11,12)], n=5 )
```


Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

```{r}
library(corrplot)
```
#correlation structure
```{r}
cij <- cor(candy)
corrplot(cij)
```
##Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
Fruity and Chocolate

##Q23. Similarly, what two variables are most positively correlated?
Answer = Chocolate and Bar



#PCA using prcomp
#need to scale the data and use summary to see it
Side-note: Feel free to examine what happens if you leave this argument out (i.e. use the default scale=FALSE). Then examine the summary(pca) and pca$rotation[,1] component and see that it is dominated by winpercent (which is after all measured on a very different scale than the other variables).
```{r}
pca <- prcomp(candy, scale =TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

#can make a nicer plot with ggplot. need a input data.frame that includes a seperate column for each of the aesthetics you would like displayed. To accomplish this we make a new data.frame here that contains our PCA results with all the rest of our candy data. We will then use this for making plots below
cbind is combining candy df and pca 1 -3 
```{r}
my_data <- cbind(candy, pca$x[,1:3])
my_data
```


```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
#library(plotly)
```

```{r}
#ggplotly(p)

```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


#Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Fruity, hard and pluribus

End
